% !TEX root = ../main.tex

\section{Introduction}

\subsection{Multi-stage Programming and MetaOCaml}

% \subsubsection{多段階計算とは何か?}

\AI{前の版は，まだ MetaOCaml specific な話と，MSP 一般の話が交互に出てきて
わかりづらかったと思います．}

Multi-stage programming makes it easier for programmers to implement
generation and execution of code at run time by providing language
constructs for composing and running pieces of code as first-class
values.  A promising application of multi-stage programming is
(run-time) code specialization, which is to generate program code
specialized to partial inputs to the program and such applications are
studied in the literature~\cite{taha2007gentle}  \AI{We should list
applications of multi-stage programming, not necessarily of MetaOCaml.}

MetaOCaml~\cite{calcagno2003implementing,oleg2014} is an extension of
OCaml\footnote{\url{http://ocaml.org}} with constructs for multi-stage
programming, including brackets, escape, run, and cross-stage
persistence (CSP) .  Brackets, written \verb|.< e >.| in MetaOCaml,
makes a value that represents the code (or abstract syntax tree) of
expression \verb|e|; and escape, written \verb|.~ e| and is supposed
to appear inside brackets, evaluates \verb|e| to a code value and
expands the code into the surrounding code.
For example,
\begin{verbatim}
   let plusone = .< fun x -> x + 1 >. in .< .~plusone 2 >.
\end{verbatim}
evaluates to \verb|.< (fun x -> x + 1) 2 >.|.  
Run, written \verb|run e|, evaluates \verb|e| to a code value
and executes it.  For example,
\begin{verbatim}
   run (let plusone = .< fun x -> x + 1 >. in .< .~plusone 2 >.)
\end{verbatim}
yields \verb|3| (here, escape connects tighter than application).
Finally, CSP is a primitive to embed values (not necessarily code
values) into a code value.  In MetaOCaml, CSP is implicitly applied to
occurrences of variables defined outside brackets.  For example,
\begin{verbatim}
   let plusone = fun x -> x + 1 in
   let y = plusone 2 in  .< y * 2 >.
\end{verbatim}
evaluates to \verb|.< 3 * 2 >.|, not \verb|.< plusone 2 * 2 >.|.
This is because the variable \verb|y| is bound to an integer \verb|3|
before \verb|.<y>.| is evaluated and the value of \verb|y| is embedded
into the code value.  In MetaOCaml, CSP can be applied to (variables of) any type.  CSP is practically important because it enables us to use library functions in brackets, e.g., \verb|.<List.combine [1;2] ['a';'b']>.|.

% \subsubsection{多段階計算のメリットをべき乗の例を用いて説明する}

% As we mentioned earlier, a main application of multi-stage
% calculi is program optimization. 
% A famous example is power functions.
% Usual power functions take two arguments, which are the base and the exponent.
% We can make a specialized power function for given exponents with multi-stage programming and 
% optimize them by unrolling a loop in functions for given exponents.
% The optimized power function for the exponent of \verb|3| looks like \verb|power3 = fun x -> x * x * x|.
% \verb|power3| is faster than ordinary \verb|power| function because it contains no loop.
% Multi-stage programming can optimize functions which take more than two arguments 
% by generating a code fragment optimized to a given argument.

% 多段階計算の型理論の既存研究の紹介

MetaOCaml is also equipped with a type system for safe code generation
and execution.  The notion of code types is introduced to prevent code
values that represent ill-typed expressions from being generated.  For
example,
\begin{verbatim}
   let plusone = .< fun x -> x + 1 >. in .< .~plusone "foo" >.
\end{verbatim}
is rejected by the MetaOCaml typechecker as \verb|plusone| is given
type \verb|(int -> int)| \verb|code| but it is used in the code context that
requires a function from \verb|string| is expected.  Ensuring safety
for \verb|run| is more challenging because basic code types do not
guarantee code values to be closed.  Taha and
Nielsen~\cite{taha2003environment} introduced the notion of
environment classifiers to address the problem and developed a type
system to ensure not only type-safe composition but also type-safe
execution of code values with a type soundness theorem.

However, the type system, which is based on Hindley--Milner
polymorphism~\cite{}, is not strong enough to guarantee invariants
beyond simple types.  For example, Kiselyov~\cite{} demonstrates
specialization of vector/matrix computation with respect to the sizes
of vectors and matrices but such specialized functions can be applied
to vectors and matrices of different sizes.

% Multi-stage programming can also optimize vector calculation.
% For example, \verb|vadd| function, which takes two vectors and return the sum of them,
% is implemented with a loop in many cases.
% We can unroll \verb|vadd| function for a given vector length and optimize it with multi-stage programming.

% % 多段階計算で生成したコードは特殊化されているが故に問題も多い

% However, there is a sever problem in functions which are optimized by multi-stage programming.
% Although unrolled \verb|vadd| function is optimized, we cannot use it for different vector length.
% For example, when you optimize for the length of 5, you shouldn't use it for vectors of 3 lengths.
% Otherwise, we will get a exception.
% This problem is serious but existing type systems for multi-stage calculi cannot prevent it.

\subsection{Multi-stage Programming with Dependent Types}

% 依存型とは何か?

One natural idea to address this problem is the introduction of dependent types
to express the size of data structure in static types~\cite{Xi98}.
% Dependent types are types which are dependent on values.
% We can use dependent types for securer programming like Xi and Pfenning
For example, we could declare vector types indexed by the size of
vectors as follows.
\begin{verbatim}
    Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor that takes an integer (which
represents the length of vectors): for example, \verb|Vector 3| is the
type for vectors whose lengths are 3.  Then, our hope is to specialize
vector/matrix functions with respect to their size and get a piece of
function code, whose type respects the given size, \emph{provided at
  specilization time}.  For example, we would like to specialize a
function to add two vectors with respect to the size of vectors, that
is, to implement a code generator that takes an integer $n$ as an
input and generates a piece of function code of type
\verb|(Vector |$n$\verb| -> Vector|$n$\verb| -> Vector |$n$\verb|) code|.

% As we pointed out in the above section,
% functions optimized with multi-stage programming can take only restricted values as arguments.
% When you optimize \verb|vadd| function for the length of 5, you should it only for 5 length vectors.
% We introduce dependent types into a multi-stage calculus
% so that the type system can guarantee optimized functions are used properly.
% Although there is a trial to combine multi-stage programming and dependent type by Brady and Hammond~\cite{brady2006dependently},
% they didn't give a formal definition and properties for the calculus.

In this paper, we develop a new multi-stage calculus \LMD by extending
the existing multi-stage calculus \LTP\cite{Hanada2014}---which
formalizes the four constructs of MetaOCaml in a way slightly
different from MetaOCaml---with dependent types and study its
properties.  We base our work on \LTP because its type system and
semantics are arguably simpler than
\(\lambda^\alpha\)~\cite{taha2003environment}, which is more faithful
to the design of MetaOCaml.  Our technical contribution are summarized
as follows:
\begin{itemize}
\item We give the formal definition of \LMD with its syntax, 
  type system and two kinds of reduction---full and staged reduction.
\item We show preservation, strong normalization, and confluence for
  full reduction; and show unique decomposition (and progress as its
  corollary) for staged reduction, which formalizes program execution.
\end{itemize}
The combination of multi-stage programming and dependent types has
been discussed by Brady and Hammond~\cite{brady2006dependently} but,
to our knowledge, our work is a first formal calculus of dependently
typed multi-stage programming.

\AI{Comments on Appendix required.}

\subsection{Organization of the Paper}

The organization of this paper is as follows.
Section~\ref{sec:informal-overview} gives an informal overview of
\LMD.  Section~\ref{sec:formal-definition} defines \LMD and
Section~\ref{sec:properties} shows properties of \LMD.  Section
\ref{sec:related-work} discusses related work and Section
\ref{sec:conclusion} concludes the paper with discussion of future
work.
